# Chapter 4 Show the Right Number
## Grouping & Faceting
```{r}
library(tidyverse)
library(gapminder)

p <- ggplot(data = gapminder, aes(x = year, y = gdpPercap))
# group
p + geom_line(aes(group = country))
# facet
p + geom_line(aes(group = country)) + facet_wrap(~continent)
# clean up
p + geom_line(aes(group = country), color = "grey70") +
    geom_smooth(size = 1.1, method = "loess", se = FALSE) +
    facet_wrap(~continent, ncol = 5) +
    scale_y_log10(labels = scales::dollar) +
    labs(x = 'Year', y = 'GDP per capita', title = 'GDP per capita on Five Continents')
ggsave("figures/Ch04_gdp_by_year_continent.pdf")    

```
## Facet Grid
```{r facet_grid}
library(tidyverse)
library(socviz)
glimpse(gss_sm)
p <- ggplot(gss_sm, aes(x = age, y = childs))
p + geom_point(alpha = 0.7) +
    facet_grid(sex ~ race) + 
    geom_smooth()
ggsave('figures/Ch04_childs_vs_age.pdf')
```
## Transform
```{r transform}
library(tidyverse)
library(socviz)
p <- ggplot(data = gss_sm, mapping = aes(x = bigregion))
p + geom_bar(mapping = aes(y = ..prop.., group = 1)) 

    # The geom_ function does a lot of work in the background.

table(gss_sm$religion)
p  <- ggplot(gss_sm, mapping = aes(x = religion))
p + geom_bar(position = "dodge",
             mapping = aes(y = ..prop.., group = bigregion)) +
    facet_wrap(~bigregion, nrow = 1)
```
Using ggplot directly to plot frequency plots can easily turn into a maze. The `dplyr` in tidyverse lets you transfrom the data before plotting, which makes catching errors easier.

## Histogram and Density Plots
```{r}
library(tidyverse)
glimpse(midwest)

p <- ggplot(data = midwest, mapping = aes(x = area))
p + geom_histogram(bins = 20)


oh_wi  <- c("OH", "WI")

p <- ggplot(data = subset(midwest, subset = state %in% oh_wi),
            mapping = aes(x = percollege, fill = state))
p + geom_histogram(alpha = 0.4, bins = 20) +
    labs(x = "Percollege", y = "Count")

p  <- ggplot(data = midwest, aes(x = area, fill = state, color = state))
p + geom_density(alpha = 0.3) + labs(x = "Area", y = "Density")
ggsave("figures/04_area_by_state.pdf", width = 6.5, height = 4, units ="in")
```
`..denstiy..` and `..scaled..` are available from `geom_denstiy`, which gives proportional density estimates. 
```{r}
library(tidyverse)
oh_wi  <- c("OH", "WI")
p  <- ggplot(data = subset(midwest,  subset = state %in% oh_wi),
            mapping = aes(x = area, fill = state, color = state))
p + geom_density(alpha = 0.3, mapping = (aes(y = ..scaled..)))
```
## Avoid Transformation When Necessary
What if I have a finished table with summary statistics already included or a grouped data without individual observations?
```{r}
library(socviz)

titanic
p  <- ggplot(titanic, aes(x = fate, y = percent, fill = sex))
p + geom_bar(position = "dodge", stat = "identity") + theme(legend.position = "top") + labs(x = "Fate", y = "Percent", fill = "Sex")

oecd_sum       
p  <- ggplot(data = oecd_sum, mapping = aes(x = year, y = diff, fill = hi_lo))
p + geom_col() + guides(fill = FALSE) + 
    labs(x = NULL, y = "Difference in Years", title = "The US Life Expectancy Gap", 
        subtitle = "Difference between US and OECD average life expectancies, 1960-2015", 
        caption = "Data: OECD. After a chart by Christopher Ingraham, Wahsington Post, December 27th 2017.")
ggsave('figures/04_US_vs_OECD_life.pdf', width = 6.5, height = 4, unit = 'in')
```
##Recap
* Map variables to aesthetics.
* Choose from variety of geoms.
* Adjust scales of plots.
* Selecct the right statistic.
* Facet plots by one or more variables.

