# Chapter 5 Graph Tables, Add Labels, Make Notes
This chapter starts diving into `dplyr`. Active verbs select, group, summarize, and transform data prior to `ggplot` plotting.

## Use Pipes to Summarize Data 
```{r pipeline}
library(socviz)
library(tidyverse)

rel_by_region  <- gss_sm %>%
    group_by(bigregion, religion) %>%
    summarize(N = n()) %>%
    mutate(freq = N / sum(N),
           pct = round((freq * 100), 0))
    
rel_by_region

p <- ggplot(rel_by_region, aes(x = bigregion, y = pct, fill = religion))
p + geom_col(position = "dodge2") +
    labs(x = "Region", y = "Percent", fill = "Religion") +
    theme(legend.position = "top")
```
This chart, however, is not a good chart. It is too crowded with too many bars stacked side-by-side.Facets increase legibility of `geom_col`. 

```{r geom_col with facet}
library(socviz)
library(tidyverse)

rel_by_region  <- gss_sm %>%
    group_by(bigregion, religion) %>%
    summarize(N = n()) %>%
    mutate(freq = N / sum(N),
           pct = round((freq * 100), 0))
    
rel_by_region

p <- ggplot(rel_by_region, aes(x = religion, y = pct, fill = religion))
p + geom_col(position = "dodge2") +
    labs(x = NULL, y = "Percent", fill = "Religion") +
    guides(fill = FALSE) +
    coord_flip() +
    facet_grid(~bigregion)
```
## Continuous Variables by Group or Category
Putting categorical variables on the y-axis using `coord\_flip` is a good way to save space when showing continuous variables.

```{r}
library(socviz)
library(tidyverse)
organdata %>% select(1:6) %>% sample_n(size = 10)

p <- ggplot(data = organdata, mapping = aes(x = reorder(country, donors, na.rm = TRUE), y = donors, color = world))
p + geom_boxplot() + coord_flip() + labs(x = NULL) + theme(legend.position = "top")
p + geom_jitter(position = position_jitter(width = 0.15)) +
    labs(x = NULL) + coord_flip() + theme(legend.position = "top")
```

`dplyr` can be used to construct summary tables
```{r summary by country}
by_country  <- organdata %>% group_by(consent_law, country) %>%
    summarize(donors_mean = mean(donors, na.rm = TRUE),
              donors_sd = sd(donors, na.rm = TRUE), 
              gdp_mean = mean(gdp, na.rm = TRUE), 
              health_mean = mean(health, na.rm = TRUE), 
              roads_mean = mean(roads, na.rm = TRUE), 
              cerebvas_mean = mean(cerebvas, na.rm = TRUE))

by_country  <- organdata %>% group_by(consent_law, country) %>%
    summarize_if(is.numeric, funs(mean, sd), na.rm = TRUE) %>%
    ungroup()

p  <- ggplot(data = by_country,
             mapping = aes(x = donors_mean, y = reorder(country, donors_mean),
             color = consent_law))
p + geom_point(size = 3) +
    labs(x = "Donor Procurement Rate", 
    y = "", color = "Consent Law") + 
    theme(legend.position = "top")


# Cleveland-style dotplot
p  <- ggplot(data = by_country,
             mapping = aes(x = donors_mean, y = reorder(country, donors_mean)))
p + geom_point(size = 3) +
    labs(x = "Donor Procurement Rate", y = "") + 
    facet_wrap(~ consent_law, scales = "free_y", ncol = 1)
```
without `free_y` each face will repeat all the countries. This is suitable for continuous variables but not for categorical variables.

```{r Cleveland-style dotplot}
p  <- ggplot(data = by_country,
             mapping = aes(y = donors_mean, x = reorder(country, donors_mean)))
p + geom_pointrange(mapping = aes(ymin = donors_mean - donors_sd,
                                  ymax = donors_mean + donors_sd)) +
    labs(x = "Donor Procurement Rate", y = "") + 
    coord_flip()
```
The `geom_pointrange` can include information about variance or error in the plot.
```{r Plot text labels with horizontal adjustment}
p <- ggplot(data = by_country,
            mapping = aes(x = roads_mean, y = donors_mean))

p + geom_point() + geom_text(mapping = aes(label = country), hjust = 0)
```
Fine tuning `hjust` often fails to provide optimal results because the added space will be proportionate to the label. Instead, we can use `ggrepel`.
```{r Use of ggrepel }
library(socviz)
library(ggrepel)
elections_historic %>% select(2:7)
p_title  <- "Presidential Elections: Popular & Electoral College Margins"
p_subtitle  <- "1824-2016"
p_caption  <- "Data for 2016 are provisional."
x_label <- "Winner's share of Popular Vote"
y_label <- "Winner's share of Electoral College Votes"

p <- ggplot(elections_historic, aes(x = popular_pct, y = ec_pct, label = winner_label))

p + geom_hline(yintercept = 0.5, size = 1.4, color = "gray80") + 
    geom_vline(xintercept = 0.5, size = 1.4, color = "gray80") +
    geom_point() +
    geom_text_repel() +
    scale_x_continuous(labels = scales::percent) + 
    scale_y_continuous(labels = scales::percent) + 
    labs(x = x_label, y = y_label, title = p_title, subtitle = p_subtitle, caption = p_caption) 
```
In addition to `geom_hline` and `geom_vline`, `geom_abline` can be used to draw guidlines when a guideline with an intercept and a slope is needed.

```{r Labeling according to criteria}
p <- ggplot(data = by_country,
            mapping = aes(x = gdp_mean, y = health_mean))

p + geom_point() +
    geom_text_repel(data = subset(by_country, gdp_mean > 25000),
                    mapping = aes(label = country))

p + geom_point() +
    geom_text_repel(data = subset(by_country,
                                  gdp_mean > 25000 | health_mean < 1500 | country %in% "Belgium"),
                    mapping = aes(label = country))
```
A dummy variable can be dedicated for labeling.
```{r Labeling using dummy variables}
organdata$ind  <- organdata$ccode %in% c("Ita", "Spa") & organdata$year > 1998
p <- ggplot(data = organdata,
            mapping = aes(x = roads,
                          y = donors, color = ind))

p + geom_point() + 
    geom_text_repel(data = subset(organdata, ind),
                    mapping = aes(label = ccode)) +
    guides(label = FALSE, color = FALSE)
```
## Write and Draw in the Plot Area
`annotate()` allows to comment or draw directly on the plot area.
```{r}
p <- ggplot(data = organdata, mapping = aes(x = roads, y = donors))
p + geom_point() + annotate(geom = "text", x = 99, y = 33,
                            label = "A surprisingly high \nrecovery rate", hjust = 0)
p + geom_point() +
    annotate(geom = "text", x = 156, y = 33,
                            label = "A surprisingly high \nrecovery rate", hjust = 0) +
    annotate(geom = "rect", xmin = 125, xmax = 155, ymin = 30, ymax = 35, fill = "red", alpha = 0.2)
```
## Understanding Scales, Guides, and Themes
```{r}
p <- ggplot(data = organdata, mapping = aes(x = roads, y = donors, color = world))
p + geom_point()
```
Scales and guides are closely related. 
